---
title: "Topic Modeling for the 'Personalized Medicine: Redefining Cancer Treatment' Challenge"
author: "Jens Hooge"
date: "01.08.2017"
output: html_document
  number_sections: true
  toc: true
  fig_width: 7
  fig_height: 4.5
  theme: cosmo
  highlight: tango
---

# Introduction
Topic models are probabilistic latent variable models of documents that exploit the correlations among the words and latent semantic themes” (Blei and Lafferty, 2007). The name "topics" signifies the hidden, to be estimated, variable relations (=distributions) that link words in a vocabulary and their occurrence in documents. A document is seen as a mixture of topics. This intuitive explanation of how documents can be generated is modeled as a stochastic process which is then "reversed"" (Blei and Lafferty, 2009) by machine learning techniques that return estimates of the latent variables. With these estimates it is possible to perform information retrieval or text mining tasks on a document corpus.

# Loading required libraries
In this study we will utilize R tm R package for querying and textmining of the literature evidence for each of the classes in the [Personalized Medicine: Redefining Cancer Treatment]() challenge on kaggle
```{r, warning=FALSE, message=FALSE}
library(tm) ## texmining
library(lda) ## the actual LDA model
library(LDAvis) # visualization library for LDA

library(parallel) # multi-core paralellization

library(data.table) # fread
library(Rmpfr) # harmonic mean maximization
library(ggplot2) # pretty plotting lib
library(reshape2) # reformatting lib for ggplot2

library(tsne) # low dimensional embedding
library(caret) # ml model wrapper lib, but only used for data transformation here

library(rbokeh) # pretty (interactive) plotting
```

# Helper Functions
```{r}
#' Copy arguments into env and re-bind any function's lexical scope to bindTargetEnv .
#' 
#' See http://winvector.github.io/Parallel/PExample.html for example use.
#' 
#' 
#' Used to send data along with a function in situations such as parallel execution 
#' (when the global environment would not be available).  Typically called within 
#' a function that constructs the worker function to pass to the parallel processes
#' (so we have a nice lexical closure to work with).
#' 
#' @param bindTargetEnv environment to bind to
#' @param objNames additional names to lookup in parent environment and bind
#' @param names of functions to NOT rebind the lexical environments of
bindToEnv <- function(bindTargetEnv=parent.frame(), objNames, doNotRebind=c()) {
  # Bind the values into environment
  # and switch any functions to this environment!
  for(var in objNames) {
    val <- get(var, envir=parent.frame())
    if(is.function(val) && (!(var %in% doNotRebind))) {
      # replace function's lexical environment with our target (DANGEROUS)
      environment(val) <- bindTargetEnv
    }
    # assign object to target environment, only after any possible alteration
    assign(var, val, envir=bindTargetEnv)
  }
}

startCluster <- function(cores=detectCores()) {
  cluster <- makeCluster(cores)
  return(cluster)
}

shutDownCluster <- function(cluster) {
  if(!is.null(cluster)) {
    stopCluster(cluster)
    cluster <- c()
  }
}

get.terms <- function(x) {
  index <- match(x, vocab)
  index <- index[!is.na(index)]
  rbind(as.integer(index - 1), as.integer(rep(1, length(index))))
}
```

First we will read the text data for the training and test datasets.
```{r}
## Read text data
train_txt_dump <- data.frame(text = read_lines('../data/original/training_text', skip = 1))
train_txt <- train_txt_dump %>%
  separate(text, into = c("ID", "txt"), sep = "\\|\\|")
train_txt <- train_txt %>%
  mutate(ID = as.integer(ID))

test_txt_dump <- data.frame(text = read_lines('../data/original/test_text', skip = 1))
test_txt <- test_txt_dump %>%
  separate(text, into = c("ID", "txt"), sep = "\\|\\|")
test_txt <- test_txt %>%
  mutate(ID = as.integer(ID))

stop_words <- stopwords("SMART")
docs <- gsub("[[:punct:]]", " ", docs)  # replace punctuation with space
docs <- gsub("[[:cntrl:]]", " ", docs)  # replace control characters with space
docs <- gsub("^[[:space:]]+", "", docs) # remove whitespace at beginning of documents
docs <- gsub("[[:space:]]+$", "", docs) # remove whitespace at end of documents
docs <- tolower(docs)  # force to lowercase
```

# Preprocessing
To train the LDA in the later steps, we need the word frequencies in each of those abstracts. For representative word frequencies we removed a number of problematic characters, removed punctuation, control characters, whitespaces, stopwords which belonged to the SMART stopword collection, all words with less than 4 characters and words which occurred less than 4 times in the documents. Lastly we transformed each word to lowercase.

```{r, eval=FALSE}
stop_words <- stopwords("SMART")
docs <- gsub("[[:punct:]]", " ", docs)  # replace punctuation with space
docs <- gsub("[[:cntrl:]]", " ", docs)  # replace control characters with space
docs <- gsub("^[[:space:]]+", "", docs) # remove whitespace at beginning of documents
docs <- gsub("[[:space:]]+$", "", docs) # remove whitespace at end of documents
docs <- tolower(docs)  # force to lowercase

# tokenize on space and output as a list:
doc.list <- strsplit(docs, "[[:space:]]+")

# Remove all words with less than 4 characters
doc.list <- lapply(doc.list, function(x) x[sapply(x, nchar)>3])

# compute the table of terms:
term.table <- table(unlist(doc.list))
term.table <- sort(term.table, decreasing = TRUE)

# remove terms that are stop words or occur fewer than 5 times:
del <- names(term.table) %in% stop_words | term.table < 5
term.table <- term.table[!del]
vocab <- names(term.table)

save(vocab, file="../data/derived/vocab.rda")
save(term.table, file="../data/derived/termTable.rda")
```

```{r, eval=TRUE}
load("../data/derived/vocab.rda")
load("../data/derived/termTable.rda")
```

Next we reformated the documents into the format required by the lda package.
```{r, eval=FALSE}
documents <- lapply(doc.list, get.terms)

save(documents, file="documents.rda")
```

Before we start training our LDA, we first will calculate some statistics related to the data set:
```{r}
D <- length(documents)  # number of documents
W <- length(vocab)  # number of terms in the vocab
doc.length <- sapply(documents, function(x) sum(x[2, ]))  # number of tokens per document
N <- sum(doc.length)  # total number of tokens in the data
term.frequency <- as.integer(term.table)  # frequencies of terms in the corpus
```

```{r}
df <- data.frame("Number of Documents"=D, 
                 "Number of Terms in Vocabulary"=W,
                 "Total Number of Tokens in Corpus"=N)

knitr::kable(df, digits = 0, caption = "Document Statistics")
```


